---
title: "p8105_hw6_crd2162"
author: "Caleigh Dwyer"
date: "2023-11-28"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(viridis)
library(modelr)
set.seed(1)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

theme_set(theme_minimal() + theme(legend.position = "bottom"))
```

### Problem 1: Homicide Data

In the data cleaning code below we create a `city_state` variable, change `victim_age` to numeric, modifiy victim_race to have categories white and non-white, with white as the reference category, and create a `resolution` variable indicating whether the homicide is solved. Lastly, we filtered out the following cities: Tulsa, AL; Dallas, TX; Phoenix, AZ; and Kansas City, MO; and we retained only the variables `city_state`, `resolution`, `victim_age`, `victim_sex`, and `victim_race`.

```{r q1_data_cleaning}
homicide_df = 
  read_csv("data/homicide-data.csv", na = c("", "NA", "Unknown")) |> 
  mutate(
    city_state = str_c(city, state, sep = ", "),
    victim_age = as.numeric(victim_age),
    resolution = case_when(
      disposition == "Closed without arrest" ~ 0,
      disposition == "Open/No arrest"        ~ 0,
      disposition == "Closed by arrest"      ~ 1)
  ) |> 
  filter(victim_race %in% c("White", "Black")) |> 
  filter(!(city_state %in% c("Tulsa, AL", "Dallas, TX", "Phoenix, AZ", "Kansas City, MO"))) |> 
  select(city_state, resolution, victim_age, victim_sex, victim_race)
```

Next we fit a logistic regression model using only data from Baltimore, MD. We model `resolved` as the outcome and `victim_age`, `victim_sex`, and `victim_race` as predictors. We save the output as `baltimore_glm` so that we can apply `broom::tidy` to this object and obtain the estimate and confidence interval of the adjusted odds ratio for solving homicides comparing non-white victims to white victims.

```{r q1_glm_baltimore}
baltimore_glm = 
  filter(homicide_df, city_state == "Baltimore, MD") |> 
  glm(resolution ~ victim_age + victim_sex + victim_race, family = binomial(), data = _)

baltimore_glm |> 
  broom::tidy() |> 
  mutate(
    OR = exp(estimate), 
    OR_CI_upper = exp(estimate + 1.96 * std.error),
    OR_CI_lower = exp(estimate - 1.96 * std.error)) |> 
  filter(term == "victim_sexMale") |> 
  select(OR, OR_CI_lower, OR_CI_upper) |>
  knitr::kable(digits = 3)
```

Below, by incorporating `nest()`, `map()`, and `unnest()` into the preceding Baltimore-specific code, we fit a model for each of the cities, and extract the adjusted odds ratio (and CI) for solving homicides comparing non-white victims to white victims. We show the first 5 rows of the resulting dataframe of model results.

```{r q1_glm_all_cities}
model_results = 
  homicide_df |> 
  nest(data = -city_state) |> 
  mutate(
    models = map(data, \(df) glm(resolution ~ victim_age + victim_sex + victim_race, 
                             family = binomial(), data = df)),
    tidy_models = map(models, broom::tidy)) |> 
  select(-models, -data) |> 
  unnest(cols = tidy_models) |> 
  mutate(
    OR = exp(estimate), 
    OR_CI_upper = exp(estimate + 1.96 * std.error),
    OR_CI_lower = exp(estimate - 1.96 * std.error)) |> 
  filter(term == "victim_sexMale") |> 
  select(city_state, OR, OR_CI_lower, OR_CI_upper)

model_results |>
  slice(1:5) |> 
  knitr::kable(digits = 3)
```

Below we generate a plot of the estimated ORs and CIs for each city, ordered by magnitude of the OR from smallest to largest. From this plot we see that most cities have odds ratios that are smaller than 1, suggesting that crimes with male victims have smaller odds of resolution compared to crimes with female victims after adjusting for victim age and race. This disparity is strongest in New York. In roughly half of these cities, confidence intervals are narrow and do not contain 1, suggesting a significant difference in resolution rates by sex after adjustment for victim age and race. 

```{r q1_plot}
model_results |> 
  mutate(city_state = fct_reorder(city_state, OR)) |> 
  ggplot(aes(x = city_state, y = OR)) + 
  geom_point() + 
  geom_errorbar(aes(ymin = OR_CI_lower, ymax = OR_CI_upper)) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


## Problem 2: Weather Data


The following chunk imports weather data from NOAA
```{r import_weather}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2022-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = recode(id, USW00094728 = "CentralPark_NY"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
  select(name, id, everything())
```

Let's create a linear model for maximum temperature (`tmax`) with minimum temperature (`tmin`) and precipitation levels (`prcp`) as predictors.

```{r lm_weather}
lm_weather =
  weather_df |> 
  lm(tmax ~ tmin + prcp, data = _)
```

Then, we'll find the r squared and log(b1*b2) for the weather model. Note that because the coefficient for prcp is negative, the log of the product of coefficients produces an NA. The log estimates are then merged with the r.squared data.

```{r}
lm_weather_rsq=
lm_weather |>
  broom::glance() |> 
  select(r.squared)

lm_weather_est=
  lm_weather |> 
  broom::tidy() |> 
  select(term, estimate) |> 
  filter(term != "(Intercept)") |> 
  summarize(log_coef = log(estimate[term == "tmin"] * estimate[term== "prcp"]))


weather_results=
bind_cols(lm_weather_rsq, lm_weather_est)

```


Now let's create 5000 bootstrap samples. We'll use map to write linear models for each bootstrap sample and produce r-squared and coefficients for each model. This produces a dataframe containing a dataframe for each bootstrap sample, a dataframe of models for each sample, a dataframe containing the r-squared, and a dataframe containing the estimates. We'll then unnest the estimate dataframe to calculate the log of the product of coefficients for each model. We'll then merge these datasets.

```{r}
bootstrap_results=
  weather_df |> 
  modelr::bootstrap(n = 5000) |> 
  mutate(
    models = map(strap, \(df) lm(tmax ~ tmin + prcp, data = df)),
    rsqr = map(models, broom::glance),
    est = map (models, broom::tidy)
  )
 
log_coef =
bootstrap_results |> 
 unnest(est) |> 
  summarize(log_coef = log(estimate[term == "tmin"] * estimate[term== "prcp"]))
  
bootstrap_df=
  bind_cols(bootstrap_results, log_coef) |> 
  unnest(rsqr) |> 
  select(strap, .id, models, r.squared, est, log_coef)
```

As mentioned, any model that includes a negative coefficient for prcp or tmin will yield a "NaN" result when calculating the log of the product of coefficients. The chunk below calculates how many bootstrap samples produced NaN log estimates for this reason. Out of 5000 bootstrap samples, an astonishing 3361 samples produced NaN log estimates.
```{r}
bootstrap_df |> 
  filter(log_coef == "NaN") |> 
  count(log_coef)
```

I will now plot the distribution of r squared estimates for the bootstrap samples. The distribution of r squared estimates shows that most estimates are centered around 0.92. The distribution is slightly left skewed.

```{r}
bootstrap_df |> 
  ggplot(aes(x = r.squared))+
  geom_histogram()+
  labs(
    title = "Distribution of R-squared estimates",
    x = "Estimates",
    y = "Frequency"
  )
```


Here is a distribution of the valid log estimates for the bootstrap samples. The distribution of log estimates shows that most estimates are centered around -6. The distribution of log estimates is left skewed as well, with a greater skew than the r squared estimates.

```{r}
bootstrap_df |> 
  filter(log_coef != "NaN") |> 
  ggplot(aes(x = log_coef))+
  geom_histogram()+
  labs(
    title = "Distribution of Log(B1*B2) estimates",
    x = "Estimates",
    y = "Frequency"
  )
```

Now we'll find the confidence interval for each estimate using the bootstrap samples:
```{r}
CI_rsqr = 
  bootstrap_df |> 
  summarize(
    lower_bound = quantile(r.squared, 0.025),
    upper_bound = quantile(r.squared, 0.975)) |> 
  print()


CI_logcoef =
  bootstrap_df |> 
  filter(log_coef != "NaN") |> 
  summarize(
    lower_bound = quantile(log_coef, 0.025),
    upper_bound = quantile(log_coef, 0.975)) |> 
  print()
```


## Problem 3: Birthweight data

First, we'll import the birthweight data:

```{r}
bw_raw = 
  read_csv("data/birthweight.csv")
```

Then, we'll tidy the data. The following data will be recoded to use actual values rather than dummy values: babysex, frace, malform, and mrace.

```{r}
bw_df = 
  bw_df |> 
  mutate(
    babysex =
      recode(
        babysex,
        "1" = "male",
        "2" = "female"
      ),
    frace = 
      recode(
        frace,
        "1" = "white",
        "2" = "black",
        "3" = "asian",
        "4" = "puerto rican",
        "8" = "other",
        "9" = "unknown"
      ),
  malform =
    recode(
      malform,
      "0" = "absent",
      "1" = "present"
    ),
  mrace=
    recode(
      mrace,
      "1" = "white",
      "2" = "black",
      "3" = "asian",
      "4" = "puerto rican",
      "8" = "other",
      "9" = "unknown"
    )
  ) |> 
  filter(frace != "unknown", mrace!= "unknown")

```

I will now use stepwise model building to construct a model that predicts birthweight. I first create an input model 0 that includes all the variables in our data set as predictors of bwt. Then, I'll use the step function to conduct the stepwise modeling. Then, I select the model with the lowest AIC from the stepwise output and name it bw_step.

```{r}
model_0=
  bw_df |> 
  lm(bwt ~ babysex + bhead + blength + delwt + fincome + frace + gaweeks + malform + menarche + mheight + momage + mrace + parity + pnumlbw + pnumsga + ppbmi + ppwt + smoken + wtgain, data = _) |> 
  step()

bw_step = 
  bw_df |> 
  lm(bwt ~ babysex + bhead + blength + delwt + fincome + gaweeks + 
    mheight + mrace + parity + ppwt + smoken, data = _)
```

Then, I'll plot the residuals from the bw_step model against the fitted values. Most of the points are centered and equally distributed around 0, indicating good model fit. However, there are some outliers curving up on the upper left portion of the plot. 

```{r}
plot_bw = 
  bw_df |> 
  add_predictions(bw_step) |> 
  add_residuals(bw_step) |> 
  select(pred, resid)

plot_bw |> 
  ggplot(aes(pred, resid))+
  geom_point()+
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Residuals vs Fitted Values",
    x = "Fitted Values",
    y = "Residuals"
  )
```


Now, we'll compare this model against two other models.
